Preferences for mercantilization: the role of intergenerational social mobility and beliefs in meritocracy in Chile

### 2025-06-27 advances/informe3.qmd
Up informe 3

What is the legitimate extent of market inequality in the eyes of the public? Since the early 1980s, many countries have experienced a widespread retreat from universal welfare programs and a shift toward the privatization and commodification of public goods, welfare policies, and social services [@gingrich_making_2011; @streeck_how_2016]. In Latin America, as elsewhere, neoliberal reforms reshaped welfare-state institutions by extending market logic into domains of social reproduction that were traditionally governed by the state [@ferre_welfare_2023]. This transformation reduced the role of public provision and increased the presence of private actors in core social services [@harvey_breve_2015]. Echoing Polanyi's [-@polanyi_great_1975] insight that markets constitute a distinct moral order, the institutional diffusion of market rules has fostered a corresponding moral economy: a constellation of norms and values concerning fair allocation, embedded in institutions and shaping individual subjectivities [@mau_inequality_2015; @svallforsMoralEconomyClass2006a]. Within this framework, a growing body of research examines the extent to which, and the mechanisms by which, citizens consider it fair that the allocation of services like health care, pensions, and education be governed by market-based criteria---a phenomenon known as *market justice preferences* [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015]. Understanding these preferences is crucial, as they contribute to legitimizing economic inequality by framing it as the fair result of individual responsibility and limited state intervention [@mau_inequality_2015].

@@

Market justice preferences are shaped not only by objective socioeconomic conditions but also by popular beliefs about inequality. Among these, meritocracy is a key normative principle underpinning market-based distributive preferences [@mau_inequality_2015]. It frames inequality as inevitable but justifiable through effort and talent [@young_rise_1958; @davis_principles_2001]. Studies show that individuals with stronger meritocratic beliefs tend to perceive less inequality and legitimize it by attributing economic differences to personal achievement [@mijs_paradox_2019; @wilson_role_2003; @batruch_belief_2023]. In highly unequal societies where access to services is largely governed by market logic, such beliefs play a critical role in normalizing inequality. Recent evidence from Chile shows that students who believe effort and talent are rewarded in their country express stronger preferences for market-based access to healthcare, pensions, and education [@castillo_socialization_2024].

@@

The empirical study of market justice preferences has employed various strategies to capture how individuals assess inequalities arising from market allocation. A common approach evaluates whether people find it fair that access to essential social services—such as healthcare, education, or pensions—depends on income. This builds on foundational work by Kluegel and Smith [-@kluegel_legitimation_1999], who examined normative justifications for capitalist systems. Subsequent studies have extended this logic to a wider set of welfare goods. For example, Immergut and Schneider [-@immergut_it_2020] and von dem Knesebeck et al. [-@vondemknesebeck_are_2016] explore whether respondents believe it is just that higher-income individuals receive better healthcare. Similarly, Lee and Stacey [-@lee_fairness_2023] apply this framework to education. Complementing these efforts, more recent research has introduced composite indicators to capture broader orientations toward market-based allocation. Lindh [-@lindh_public_2015], for instance, constructs an index averaging support for income-based access to healthcare and education in comparative perspective, while Castillo et al. [-@castillo_socialization_2024] propose a single-item measure covering health, education, and pensions in Chile. These instruments seek to gauge to what extent individuals view market-generated inequalities as legitimate. Taken together, they capture two core dimensions of market distribution: the role of economic resources as a key determinant of outcomes, and the framing of services as tradable commodities that can be bought and sold according to ability to pay [-@lindh_public_2015].

Comparative empirical research has identified several individual-level factors that influence support for market justice. Individuals in more advantaged socioeconomic positions—those with higher income, education, and occupational status—are consistently more likely to endorse market-based distributive principles [@koos_moral_2019; @lindh_public_2015; @svallfors_political_2007]. For example, Lindh [-@lindh_public_2015] finds that individuals from the service class are more likely to support market-based access to healthcare and education than skilled and unskilled workers across 17 relatively affluent countries. In a comparative analysis, Svallfors [-@svallfors_political_2007] observes that this expected class pattern appears clearly only in Sweden, where support for private education and healthcare varies systematically by class. Busemeyer [-@busemeyer_skills_2014] similarly shows that support for private education is stronger among high-income groups, while Immergut and Schneider [-@immergut_it_2020] and von dem Knesebeck et al. [-@vondemknesebeck_are_2016] report comparable findings for healthcare, suggesting that wealthier individuals perceive private provision as a means of maintaining relative advantage. In Chile, Otero and Mendoza [-@otero_power_2024] show that individuals with higher income and university education express stronger support for market allocation in healthcare, education, and pensions. Beliefs about inequality and political orientation also matter. In Chile, Castillo et al. [-@castillo_socialization_2024] show that individuals who have strong meritocratic beliefs are more likely to support market-based distribution, while Lee and Stacy [-@lee_fairness_2023] in Australia suggest that these preferences are also greater as people lean toward economic conservatism. In this sense, market justice preferences are shaped by the interplay between structural position and normative reasoning.

@@

These dimensions of meritocracy reflect how respondents perceive society's distributive logic, regardless of whether they endorse meritocratic principles. This distinction aligns with recent findings indicating that individuals distinguish between how merit is perceived in society and how it should ideally operate, which in turn shapes their preferences for redistribution and justice [@tejero-peregrina_perceived_2025]. Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity. Prior studies have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes [@castillo_meritocracia_2019; @castillo_socialization_2024].

@@

@@

This study draws on data from the Chilean Longitudinal Social Survey (ELSOC) of the Center for Social Conflict and Cohesion Studies (COES). The survey is an annual panel collected from 2016 to 2023, comprising two independent samples (original and refreshment), featuring permanent and rotating questionnaire modules. It evaluates how individuals think, feel, and behave regarding conflict and social cohesion in Chile. The sampling design is complex, probabilistic, clustered, multistage, and stratified by city size, targeting men and women aged 18--75 who habitually reside in private dwellings in urban areas spanning 40 cities (92 municipalities, 13 regions). This analysis focuses on 2023, the most recent wave, which includes 1,737 respondents in the original sample and 989 in the refreshment sample. Further details on sampling, attrition, and weighting can be found at https://coes.cl/encuesta-panel/, and the dataset is publicly available at https://dataverse.harvard.edu/dataverse/elsoc.

>>>>

What is the legitimate extent of market inequality in the eyes of the public? Since the early 1980s, many countries have experienced a widespread retreat from universal welfare programs and a shift toward the privatization and commodification of public goods, welfare policies, and social services [@gingrich_making_2011; @streeck_how_2016]. In Latin America, as elsewhere, neoliberal reforms reshaped welfare-state institutions by extending market logic into domains of social reproduction that were traditionally governed by the state [@ferre_welfare_2023]. This transformation reduced the role of public provision and increased the presence of private actors in core social services [@ferre_welfare_2023]. Echoing Polanyi's [-@polanyi_great_1975] insight that markets constitute a distinct moral order, the institutional diffusion of market rules has fostered a corresponding moral economy: a constellation of norms and values concerning fair allocation, embedded in institutions and shaping individual subjectivities [@mau_inequality_2015; @svallforsMoralEconomyClass2006a]. Within this framework, a growing body of research examines the extent to which, and the mechanisms by which, citizens consider it fair that the allocation of services like health care, pensions, and education be governed by market-based criteria---a phenomenon known as *market justice preferences* [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @castillo_perceptions_2025; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015]. Understanding these preferences is crucial, as they contribute to legitimizing economic inequality by framing it as the fair result of individual responsibility and limited state intervention [@mau_inequality_2015].

@@

Market justice preferences are shaped not only by objective socioeconomic conditions but also by popular beliefs about inequality. Among these, meritocracy is a key normative principle underpinning market-based distributive preferences [@mau_inequality_2015]. Studies show that individuals with stronger meritocratic beliefs tend to perceive less inequality and legitimize it by attributing economic differences to personal achievement [@mijs_paradox_2019; @wilson_role_2003; @batruch_belief_2023]. In highly unequal societies where access to services is largely governed by market logic, such beliefs play a critical role in normalizing inequality. Recent evidence from Chile shows that students who believe effort and talent are rewarded in their country express stronger preferences for market-based access to healthcare, pensions, and education [@castillo_socialization_2024].

@@

The empirical study of market justice preferences has employed various strategies to capture how individuals assess inequalities arising from market allocation. A common approach evaluates whether people find it fair that access to essential social services—such as healthcare, education, or pensions—depends on income. This builds on foundational work by Kluegel and Smith [-@kluegel_legitimation_1999], who examined normative justifications for capitalist systems. Subsequent studies have extended this logic to a wider set of welfare goods. For example, Immergut and Schneider [-@immergut_it_2020] and von dem Knesebeck et al. [-@vondemknesebeck_are_2016] explore whether respondents believe it is just that higher-income individuals receive better healthcare. Similarly, Lee and Stacey [-@lee_fairness_2023] and @castillo_changes_2025 apply this framework to education. Complementing these efforts, more recent research has introduced composite indicators to capture broader orientations toward market-based allocation. Lindh [-@lindh_public_2015], for instance, constructs an index averaging support for income-based access to healthcare and education in comparative perspective, while Castillo et al. [-@castillo_socialization_2024] propose a single-item measure covering health, education, and pensions in Chile. These instruments seek to gauge to what extent individuals view market-generated inequalities as legitimate. Taken together, they capture two core dimensions of market distribution: the role of economic resources as a key determinant of outcomes, and the framing of services as tradable commodities that can be bought and sold according to ability to pay [@lindh_public_2015].

Comparative empirical research has identified several individual-level factors that influence support for market justice. Individuals in more advantaged socioeconomic positions—those with higher income, education, and occupational status—are consistently more likely to endorse market-based distributive principles [@koos_moral_2019; @lindh_public_2015; @svallfors_political_2007]. For example, Lindh [-@lindh_public_2015] finds that individuals from the service class are more likely to support market-based access to healthcare and education than skilled and unskilled workers across 17 relatively affluent countries. In a comparative analysis, Svallfors [-@svallfors_political_2007] observes that this expected class pattern appears clearly only in Sweden, where support for private education and healthcare varies systematically by class. Busemeyer [-@busemeyer_skills_2014] similarly shows that support for private education is stronger among high-income groups, while Immergut and Schneider [-@immergut_it_2020] and von dem Knesebeck et al. [-@vondemknesebeck_are_2016] report comparable findings for healthcare, suggesting that wealthier individuals perceive private provision as a means of maintaining relative advantage. In Chile, Otero and Mendoza [-@otero_power_2024] show that individuals with higher income and university education express stronger support for market allocation in healthcare, education, and pensions. Beliefs about inequality and political orientation also matter. Castillo et al. [-@castillo_socialization_2024] and @castillo_perceptions_2025 show that individuals who have strong meritocratic beliefs are more likely to support market-based distribution in Chile, while Lee and Stacy [-@lee_fairness_2023] in Australia suggest that these preferences are also greater as people lean toward economic conservatism. In this sense, market justice preferences are shaped by the interplay between structural position and normative reasoning.

@@

These dimensions of meritocracy reflect how respondents perceive society's distributive logic, regardless of whether they endorse meritocratic principles. This distinction aligns with recent findings indicating that individuals distinguish between how merit is perceived in society and how it should ideally operate, which in turn shapes their preferences for redistribution and justice [@tejero-peregrina_perceived_2025]. Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity. Prior studies in Chile have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes in healthcare, education and pensions [@castillo_perceptions_2025; @castillo_socialization_2024].

@@

Chile offers a compelling case for examining individual preferences for market justice in the access to core social services. Since the early 1990s, the country experienced a long period of sustained economic growth, with per capita GDP increasing by nearly 4% annually until the mid-2010s [@llorca-jana_historia_2021]. This growth brought notable improvements in living standards, including a strong decline in income poverty—from 38.6% in 1990 to 6.5% in 2022 [@mideplan_evolucion_2017]. However, this success story has been tempered by growing concerns over the unequal distribution of its benefits. In recent years, economic growth has slowed—exacerbated by the global impact of the COVID-19 pandemic—and structural inequalities have become more visible [@barozet_clases_2021]. Despite its economic achievements, Chile remains one of the most unequal countries in Latin America and the OCDE. The poorest 50% of the population captures just 10% of total income and holds negative net wealth, while the wealthiest 1% receives nearly 27% of all income and controls almost half (49.6%) of the country’s wealth [@chancel_world_2022], a figure that has seen little change since the 1990s [@flores_top_2020]. These stark disparities extend beyond income, reflecting deep inequalities in access to essential social services [@pnud_desiguales_2017].

A significant share of Chile’s inequality stems from neoliberal reforms that privatized and commodified key domains of social reproduction [@arrizabalo_milagro_1995; @ferre_welfare_2023]. Introduced during the military dictatorship (1973–1989) and expanded under democratic governments, these reforms embedded market logic into public services through concessions, subsidies, demand-side vouchers, and pro-private regulatory frameworks [@madariaga_three_2020; @boccardo_30_2020]. This "crowded-out" welfare model benefits higher-income groups, leaving lower-income individuals to rely on limited public options. Scholars argue that this neoliberal shift repurposed the state to create niches of capitalist accumulation -from water to education-, giving rise to a model of “public-service capitalism” heavily reliant on state funding [@boccardo_30_2020]. In healthcare, around 79% of the population is covered by the public insurer FONASA, while 15% are enrolled in private ISAPREs, which offer faster and higher-quality services to wealthier groups [@mideplan_estadisticas_2024]. The pension system is based on individual capitalization, with mandatory contributions managed by private administrators investing in the financial market, currently involving 11 million contributors through private fund administrators (AFPs), yet 27% of the labor force remains outside the system due to informality [@superintendenciadepensiones_estadisticas_2024]. Education is similarly stratified: only 30.6% of students attend fully public schools, while 54% are in voucher-funded private institutions and 9.3% in fully private schools that largely serve affluent families [@ministeriodeeducacion_resumen_2023]. This arrangement has created a structurally segmented system, where access and quality of services depend on ability to pay, reinforced by state subsidies to private providers [@pnud_desiguales_2017].

Over recent decades, Chile has undergone a significant transformation in its class structure, shaped by sustained economic growth, the expansion of the service sector, the massification of education, and major changes in the role of the state and social policy [@ruiz_chilenos_2014]. These changes contributed to the growth of the middle sectors, a phenomena labelled as “mesocratrization” of society [@espinoza_estratificacion_2013]. However, this expansion did not reduce structural inequalities nor facilitate upward mobility into elite positions [@espinoza_estratificacion_2013; @lopez-roldan_comparative_2021]. While Chile shows high absolute mobility, most flows occur between adjacent classes—particularly from working classes into lower service-class positions—while access to upper classes remains limited, indicating strong elite closure and reproduction [@espinoza_estratificacion_2013; @lopez-roldan_comparative_2021; @torche_unequal_2005; @espinoza_movilidad_2014; @perez-ahumada_clases_2019]. Though often described as “unequal but fluid”, recent evidence points to growing rigidity in relative mobility, especially after the economic slowdown, increased household debt, labor market precarity, and the COVID-19 crisis [@barozet_clases_2021]. However, the majority of the population identifies themselves as members of the middle class [@castillo_todos_2013]. These dynamics have fueled a legitimacy crisis: meritocracy is increasingly questioned, and many perceive that advancement depends more on informal networks than on effort or earned credentials [@barozet_clases_2021]. The combination of constrained mobility, persistent inequality, and concentrated opportunity has eroded expectations of upward mobility and intensified social discontent [@pnud_desiguales_2017].

Despite recurrent social unrest, Chile presents a paradoxical coexistence between strong conflict over inequality and widespread public legitimation of it. The October 2019 “social outburst”, a period of mass protests and severe political repression throughout the country,  crystallized discontent over the privatization and commodification of social services, alongside a crisis of political legitimacy [@somma_no_2021]. A survey during the protests identified pensions, healthcare, and education as the top demands [@nucleodesociologiacontingente_informe_2020], reflecting dissatisfaction with Chile’s stratified welfare regime. Yet, this unrest coexists with strong meritocratic beliefs and high tolerance for income inequality. Castillo [-@castillo_legitimacy_2011] shows that wage gaps between occupations are consistently justified, and the larger the perceived gap, the greater its legitimation, an effect amplified by Chile’s structural inequality. Moreover, perceptions of unfairness due to non-meritocratic factors often reinforce belief in effort and talent as legitimate bases for success [@castillo_multidimensional_2023], which are associated with lower perceived injustice [@castillo_meritocracia_2019]. Mac-Clure et al. [@mac-clure_justicia_2024] find that low and lower-middle status individuals are less likely than upper-status groups to view educational income differences as unfair. Qualitative research further reveals the internalization of the enterprising self [@mau_inequality_2015], understood as a way in which the subjectivization process arises in line with positioning the self in the market, generating its own market value and subjecting it to competition. Canales et al. [-@canalesceron_sujeto_2021] describes how families navigate school choice as investment decisions, while Panes [-@panes_criticas_2020] finds that many workers frame pension contributions as individual investments aimed at maximizing future returns. These representations suggest that market-based moral orientations are deeply embedded in Chilean subjectivities, offering fertile ground to study how support for market justice varies across welfare domains.

Although still limited, recent research on market justice preferences in Chile indicates that support for these beliefs has increased in recent years across domains such as healthcare, pensions, and education, and that both objective and subjective dimensions of inequality shape them. Castillo et al. [-@castillo_perceptions_2025] show that agreement with the notion that it is fair for higher-income individuals to access better services was relatively low in 2016, but rose significantly by 2023—especially in the pension domain. Otero and Mendoza [-@otero_power_2024] provide evidence that social class, measured using the EGP scheme, influences these preferences: individuals in lower or subordinate class positions are less supportive of market-based access to welfare services compared to those in higher or privileged positions. Additionally, greater diversity in social class networks is associated with lower support for market justice. This is particularly relevant in the Chilean context, where one of the main determinants of class-diverse networks is intergenerational social mobility, which exposes individuals to relationships with people from different class backgrounds [@otero_lives_2022]. Regarding subjective factors, Castillo et al. [@castillo_perceptions_2025; @castillo_socialization_2024] find that individuals who endorse meritocratic beliefs—such as the idea that effort and talent are rewarded in Chile—are more likely to support market justice across welfare domains.

@@

This study draws on data from the Chilean Longitudinal Social Survey (ELSOC) of the Center for Social Conflict and Cohesion Studies (COES). The survey is an annual panel collected from 2016 to 2023, comprising two independent samples (original and refreshment), featuring permanent and rotating questionnaire modules. It evaluates how individuals think, feel, and behave regarding conflict and social cohesion in Chile. The sampling design is complex, probabilistic, clustered, multistage, and stratified by city size, targeting men and women aged 18--75 who habitually reside in private dwellings in urban areas spanning 40 cities (92 municipalities, 13 regions). This analysis focuses on 2023, the most recent wave, which includes 1,737 respondents in the original sample and 989 in the refreshment sample. Further details on sampling, attrition, and weighting can be found at [https://coes.cl/encuesta-panel/](https://coes.cl/encuesta-panel/), and the dataset is publicly available at [https://dataverse.harvard.edu/dataverse/elsoc](https://dataverse.harvard.edu/dataverse/elsoc).
